Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-06 survey/survey.tex
before merge

Bijections on \fin{n} have a specific name: permutations. As is
well-known permutations can be generated by sequential compositions of
transpositions. Thus we can create a small language for writing
permutations on $\fin{n}$ as
\[\begin{array}{rcl}
p^n &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

For convenience, we write $[2^n]$ for the finite set representing
$n$-bit words with the canonical ordering for binary numbers. Thus
when $n=3$, the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which
correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
Although this language appears weak, it is universal for boolean
combinational circuits. To see this, we express, boolean negation
\emph{not} as a permutation $[2^1] \rightarrow [2^1]$, controlled not
(also known as \emph{cnot}) as a permutation
$[2^2] \rightarrow [2^2]$, and controlled controlled-not (also known
as \emph{toffoli}) as a permutation $[2^3] \rightarrow [2^3]$:
\[\begin{array}{rcl}
\mathit{not} &=& \mathit{swap}~0~1 \\
\mathit{cnot} &=& \mathit{swap}~2~3 \\
\mathit{toffoli} &=& \mathit{swap}~6~7
\end{array}\]
For example, \emph{toffoli} gate negates the third bit (the target
bit) if both the first two bits (the control bits) are 1, i.e., it
swaps $110$ and $111$. Universality follows since any combinational
circuit can be realized using compositions of \emph{toffoli} gates.

@@

\amr{***HERE***; operational semantics; It feels like I should be
  writing about how the combinators (both for spaces and for
  'permutations') work at a lower level. Did we ever write that up
  somewhere?  We have lots of Agda code which 'does it', nicely, but
  somehow I can't remember an English prose version of it. I am not
  sure how long we want to make it.  For example, do we talk about the
  'matrix' representation of combinators-for-permutations-on-[n], and
  how these make things like 0+a=a and associativity obvious by
  construction?
  Let’s have a geometric representation of spaces:
- A + B means putting A next to B
- A * B means putting A and B on two perpendicular axes and filling the square (volume) between them
All the combinators would indeed have a natural action and if we have
a few nice pictures it would be great. What do you think? }
\jc{You are correct, I don't think we have any English prose which
explains the operations of our combinators. The `matrix' representation
makes sense for the untyped version, but is not necessarily adequate for
the typed version, as it becomes quite an indirect way to view it, since you
first have to assign numbers to each 'path' through a type as a proxy.
Though if we want to have path-indexed instead of Fin-indexed matrices,
I would be fine. So we could indeed still draw those pictures.  However,
this would lead associativity of * to be an equation on two 3D matrices...}

@@

>>>>

Bijections on \fin{n} have a specific name: permutations. As is
well-known, permutations can be generated by sequential compositions of
transpositions. Thus we can create a small language for writing
permutations on $\fin{n}$ as
\[\begin{array}{rcl}
p^n &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

For convenience, we write $[2^n]$ for the finite set representing
$n$-bit words with the canonical ordering for binary numbers. Thus
when $n=3$, the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which
correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
Although this language appears weak, it is universal for boolean
combinational circuits. To see this, we express, boolean negation
\emph{not} as a permutation $[2^1] \rightarrow [2^1]$, controlled not
(also known as \emph{cnot}) as a permutation
$[2^2] \rightarrow [2^2]$, and controlled controlled-not (also known
as \emph{toffoli}) as a permutation $[2^3] \rightarrow [2^3]$:
\[\begin{array}{rcl}
\mathit{not} &=& \mathit{swap}~0~1 \\
\mathit{cnot} &=& \mathit{swap}~2~3 \\
\mathit{toffoli} &=& \mathit{swap}~6~7
\end{array}\]
For example, the \emph{toffoli} gate negates the third bit (the target
bit) if both the first two bits (the control bits) are 1, i.e., it
swaps $110$ and $111$. Universality follows since any combinational
circuit can be realized using compositions of \emph{toffoli} gates.

@@

\amr{***HERE***; operational semantics; It feels like I should be
  writing about how the combinators (both for spaces and for
  'permutations') work at a lower level. Did we ever write that up
  somewhere?  We have lots of Agda code which 'does it', nicely, but
  somehow I can't remember an English prose version of it. I am not
  sure how long we want to make it.  For example, do we talk about the
  'matrix' representation of combinators-for-permutations-on-[n], and
  how these make things like 0+a=a and associativity obvious by
  construction?
  Let’s have a geometric representation of spaces:
- A + B means putting A next to B
- A * B means putting A and B on two perpendicular axes and filling the square (volume) between them
All the combinators would indeed have a natural action and if we have
a few nice pictures it would be great. What do you think? }

@@

Must mention infinite sets since we promised to do so in Sec 2
